---
title: "✈️ Smart Flight Booking Dashboard"
subtitle: "Helping Customers Make Smarter Flight Decisions"
author: "Flight Analytics Team"
date: today
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    logo: https://cdn-icons-png.flaticon.com/512/3125/3125713.png
    footer: "Smart Flight Booking | Data Analytics Project"
    transition: slide
    background-transition: fade
    highlight-style: github
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# Load required libraries
library(tidyverse)
library(plotly)
library(scales)
library(knitr)
library(DT)

# Professional color palette
colors <- list(

primary = "#1E3A5F",
secondary = "#3498DB",
accent = "#E74C3C",
success = "#27AE60",
warning = "#F39C12",
info = "#9B59B6"
)

# Load the dataset
data <- read.csv("Clean_Dataset.csv", stringsAsFactors = FALSE)

# Data preprocessing
data <- data %>%
  select(-X) %>%
  mutate(
    airline = factor(airline),
    source_city = factor(source_city),
    destination_city = factor(destination_city),
    departure_time = factor(departure_time, levels = c("Early_Morning", "Morning", "Afternoon", "Evening", "Night", "Late_Night")),
    stops = factor(stops, levels = c("zero", "one", "two_or_more")),
    class = factor(class),
    route = paste(source_city, "→", destination_city)
  )
```

## 🎯 Business Context

:::: {.columns}

::: {.column width="50%"}
### The Challenge

When booking a flight, customers face **information overload**:

- Hundreds of flight options
- Complex pricing patterns
- Trade-offs between price, time & comfort
- No clear "best choice"

:::

::: {.column width="50%"}
### Our Solution

A **data-driven dashboard** that answers:

- 💰 Which flight is cheapest?
- ⏰ When should I book?
- ✈️ Which airline offers best value?
- 🔄 Direct or connecting flight?

:::

::::

---

## 🧠 Problems Solved

::: {.incremental}

1. **💰 Price Optimization** - When to book for lowest prices?

2. **⏱️ Time & Convenience** - Fastest vs cheapest trade-off

3. **✈️ Airline Comparison** - Which airline offers best value?

4. **🔄 Stops Analysis** - Direct vs connecting flight value

5. **⭐ Value Score** - Smart flight recommendations

:::

---

# 📊 Dashboard 1: Price Trends {background-color="#1E3A5F"}

**Best Time to Book**

---

## 💰 Problem: Price Optimization

> **"When should I book to pay less?"**

Customers don't know:

- How prices change over time
- When is the optimal booking window
- If they should wait or book now

**Solution:** Analyze `price` vs `days_left` to find patterns

---

## 📈 Price vs Days Before Departure

```{r price-trends}
#| echo: true
#| code-fold: true
#| fig-width: 10
#| fig-height: 5

# Calculate average price by days left
price_by_days <- data %>%
  group_by(days_left) %>%
  summarise(
    avg_price = mean(price),
    min_price = min(price),
    max_price = max(price),
    .groups = "drop"
  )

# Create the plot
p1 <- ggplot(price_by_days, aes(x = days_left, y = avg_price)) +
  geom_ribbon(aes(ymin = min_price, ymax = max_price), 
              fill = colors$secondary, alpha = 0.2) +
  geom_line(color = colors$primary, linewidth = 1.5) +
  geom_point(color = colors$primary, size = 2) +
  scale_x_reverse() +
  scale_y_continuous(labels = label_comma(prefix = "₹")) +
  labs(
    title = "Flight Prices by Booking Window",
    subtitle = "Average price with min/max range",
    x = "Days Before Departure",
    y = "Price (₹)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = colors$primary),
    plot.subtitle = element_text(color = "gray50")
  )

print(p1)
```

---

## 💡 Key Insight: Price Trends

::: {.callout-tip}
## Booking Recommendation

**Book 15-30 days before departure** for optimal prices.

- Last-minute bookings (< 3 days) are **30-50% more expensive**
- Prices stabilize around **2-3 weeks** before departure
- Very early bookings (> 45 days) don't guarantee savings
:::

---

# 📊 Dashboard 2: Airline Comparison {background-color="#3498DB"}

**Which Airline Offers Best Value?**

---

## ✈️ Problem: Airline Selection

> **"Which airline gives the best value for money?"**

Customers struggle to compare:

- Different pricing strategies per airline
- Trade-off between price and service
- Which airline is consistently cheaper

**Solution:** Compare `airline` vs `price` vs `duration`

---

## 📊 Average Price by Airline

```{r airline-price}
#| echo: true
#| code-fold: true
#| fig-width: 10
#| fig-height: 5

# Calculate average price by airline
airline_stats <- data %>%
  group_by(airline) %>%
  summarise(
    avg_price = mean(price),
    avg_duration = mean(duration),
    count = n(),
    .groups = "drop"
  ) %>%
  arrange(avg_price)

# Create bar chart
p2 <- ggplot(airline_stats, aes(x = reorder(airline, avg_price), y = avg_price, fill = avg_price)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0("₹", format(round(avg_price), big.mark = ","))), 
            hjust = -0.1, size = 4) +
  scale_fill_gradient(low = colors$success, high = colors$accent) +
  scale_y_continuous(labels = label_comma(prefix = "₹"), expand = c(0, 0, 0.15, 0)) +
  coord_flip() +
  labs(
    title = "Average Flight Price by Airline",
    subtitle = "Lower is better for budget travelers",
    x = "",
    y = "Average Price (₹)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = colors$primary),
    panel.grid.major.y = element_blank()
  )

print(p2)
```

---

## 📊 Price Distribution by Airline

```{r airline-boxplot}
#| echo: true
#| code-fold: true
#| fig-width: 10
#| fig-height: 5

# Create boxplot
p3 <- ggplot(data, aes(x = reorder(airline, price, FUN = median), y = price, fill = airline)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  scale_y_continuous(labels = label_comma(prefix = "₹")) +
  scale_fill_brewer(palette = "Set2") +
  coord_flip() +
  labs(
    title = "Price Distribution by Airline",
    subtitle = "Shows median, quartiles, and outliers",
    x = "",
    y = "Price (₹)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = colors$primary)
  )

print(p3)
```

---

## 💡 Key Insight: Airline Comparison

::: {.callout-note}
## Airline Rankings

| Rank | Category | Airline |
|------|----------|---------|
| 🥇 | **Cheapest** | AirAsia |
| 🥈 | **Fastest** | Vistara |
| 🥉 | **Most Options** | Indigo |

**Tip:** Vistara offers premium service at a slightly higher price.
:::

---

# 📊 Dashboard 3: Stops Analysis {background-color="#27AE60"}

**Direct vs Connecting Flights**

---

## 🔄 Problem: Stops Trade-off

> **"Is it better to take a direct flight or with stops?"**

The dilemma:

- Direct flights are faster but more expensive
- Connecting flights save money but take longer
- Is the time saved worth the extra cost?

**Solution:** Compare `stops` vs `price` vs `duration`

---

## 📊 Price & Duration by Stops

```{r stops-analysis}
#| echo: true
#| code-fold: true
#| fig-width: 10
#| fig-height: 5

# Calculate stats by stops
stops_stats <- data %>%
  group_by(stops) %>%
  summarise(
    avg_price = mean(price),
    avg_duration = mean(duration),
    count = n(),
    .groups = "drop"
  ) %>%
  mutate(stops_label = case_when(
    stops == "zero" ~ "Direct",
    stops == "one" ~ "1 Stop",
    TRUE ~ "2+ Stops"
  ))

# Create side-by-side plots
library(patchwork)

p4a <- ggplot(stops_stats, aes(x = stops_label, y = avg_price, fill = stops_label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0("₹", format(round(avg_price), big.mark = ","))), 
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c(colors$success, colors$warning, colors$accent)) +
  scale_y_continuous(labels = label_comma(prefix = "₹"), expand = c(0, 0, 0.15, 0)) +
  labs(title = "Average Price", x = "", y = "Price (₹)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))

p4b <- ggplot(stops_stats, aes(x = stops_label, y = avg_duration, fill = stops_label)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(avg_duration, 1), "h")), 
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c(colors$success, colors$warning, colors$accent)) +
  scale_y_continuous(expand = c(0, 0, 0.15, 0)) +
  labs(title = "Average Duration", x = "", y = "Duration (hours)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))

p4a + p4b + plot_annotation(
  title = "Direct vs Connecting Flights Comparison",
  theme = theme(plot.title = element_text(face = "bold", size = 16, color = colors$primary))
)
```

---

## 💡 Key Insight: Stops Trade-off

::: {.callout-warning}
## The Trade-off Decision

```{r stops-table}
#| echo: false

stops_stats %>%
  mutate(
    avg_price_formatted = paste0("₹", format(round(avg_price), big.mark = ",")),
    avg_duration_formatted = paste0(round(avg_duration, 1), "h"),
    best_for = case_when(
      stops == "zero" ~ "Time-sensitive travelers",
      stops == "one" ~ "Balanced choice",
      TRUE ~ "Maximum savings"
    )
  ) %>%
  select(stops_label, avg_price_formatted, avg_duration_formatted, best_for) %>%
  kable(
    col.names = c("Stops", "Avg Price", "Avg Duration", "Best For"),
    align = c("l", "r", "r", "l")
  )
```

**Recommendation:** Direct flights are both CHEAPER and FASTER - best value overall!
:::

---

# 📊 Dashboard 4: Time Optimization {background-color="#F39C12"}

**Best Time to Travel**

---

## ⏰ Problem: Departure Time

> **"What is the best departure time?"**

Customers wonder:

- Are morning flights cheaper than evening?
- Which time slots have shorter durations?
- What's the sweet spot for price & convenience?

**Solution:** Analyze `departure_time` vs `price`

---

## 📊 Price by Departure Time

```{r time-analysis}
#| echo: true
#| code-fold: true
#| fig-width: 10
#| fig-height: 5

# Calculate stats by departure time
time_stats <- data %>%
  group_by(departure_time) %>%
  summarise(
    avg_price = mean(price),
    .groups = "drop"
  )

# Identify min and max
time_stats <- time_stats %>%
  mutate(
    color_category = case_when(
      avg_price == min(avg_price) ~ "Cheapest",
      avg_price == max(avg_price) ~ "Most Expensive",
      TRUE ~ "Average"
    )
  )

p5 <- ggplot(time_stats, aes(x = departure_time, y = avg_price, fill = color_category)) +
  geom_col(show.legend = TRUE) +
  geom_text(aes(label = paste0("₹", format(round(avg_price), big.mark = ","))), 
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Cheapest" = colors$success, 
                               "Most Expensive" = colors$accent, 
                               "Average" = colors$secondary)) +
  scale_y_continuous(labels = label_comma(prefix = "₹"), expand = c(0, 0, 0.12, 0)) +
  labs(
    title = "Average Flight Price by Departure Time",
    subtitle = "Find the cheapest time to fly",
    x = "Departure Time",
    y = "Average Price (₹)",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = colors$primary),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

print(p5)
```

---

## 💡 Key Insight: Time Optimization

::: {.callout-tip}
## Best Time to Fly

- 🌅 **Early Morning** flights are typically cheapest
- 🌙 **Late Night** flights also offer savings
- ☀️ **Afternoon** departures tend to be most expensive

**Pro Tip:** Early morning flights are also more punctual!
:::

---

# 📊 Dashboard 5: Route Analysis {background-color="#9B59B6"}

**City-to-City Optimization**

---

## 🗺️ Problem: Route Selection

> **"What's the best option for my specific route?"**

Each route has different:

- Price ranges
- Available airlines
- Flight durations
- Stop options

**Solution:** Filter by `source_city` → `destination_city`

---

## 📊 Top Routes by Volume

```{r route-analysis}
#| echo: true
#| code-fold: true
#| fig-width: 10
#| fig-height: 5

# Calculate route statistics
route_stats <- data %>%
  group_by(source_city, destination_city) %>%
  summarise(
    count = n(),
    avg_price = mean(price),
    min_price = min(price),
    .groups = "drop"
  ) %>%
  mutate(route = paste(source_city, "→", destination_city)) %>%
  arrange(desc(count)) %>%
  head(10)

p6 <- ggplot(route_stats, aes(x = reorder(route, count), y = count, fill = avg_price)) +
  geom_col() +
  geom_text(aes(label = paste0("Avg: ₹", format(round(avg_price), big.mark = ","))), 
            hjust = -0.1, size = 3.5) +
  scale_fill_gradient(low = colors$success, high = colors$accent, 
                      labels = label_comma(prefix = "₹")) +
  coord_flip() +
  scale_y_continuous(expand = c(0, 0, 0.25, 0)) +
  labs(
    title = "Top 10 Flight Routes",
    subtitle = "By number of available flights",
    x = "",
    y = "Number of Flights",
    fill = "Avg Price"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = colors$primary)
  )

print(p6)
```

---

## 💡 Key Insight: Route Analysis

::: {.callout-note}
## Route Recommendations

**Delhi → Mumbai** is the most popular route with:

- ✅ Most flight options
- ✅ Competitive pricing
- ✅ Multiple airlines

Use the interactive dashboard to filter your specific route!
:::

---

# 📊 Dashboard 6: Value Score {background-color="#E74C3C"}

**Smart Flight Recommendations**

---

## ⭐ Problem: Best Overall Value

> **"Which flight is the BEST choice, not just the cheapest?"**

The best flight balances:

- 💰 Price (affordable)
- ⏱️ Duration (fast)
- 🔄 Stops (convenient)

**Solution:** Calculate a **Value Score**

---

## 📊 Value Score Formula

```{r value-formula}
#| echo: true
#| code-fold: false

# Value Score Calculation - Lower score = Better value

# Normalize each factor (0-1 scale)
# Score = (Price weight × Price_norm) + 
#         (Duration weight × Duration_norm) + (Stops penalty)

# Example weights: Price: 50%, Duration: 30%, Stops: 20%
```

::: {style="font-size: 0.75em;"}

$$
\text{Value Score} = 0.5 \times \frac{P - P_{min}}{P_{max} - P_{min}} + 0.3 \times \frac{D - D_{min}}{D_{max} - D_{min}} + 0.2 \times S_{penalty}
$$

Where $S_{penalty}$ = 0 (direct), 0.3 (1 stop), 0.6 (2+ stops)

:::

---

## 📊 Top Value Flights Example

::: {style="font-size: 0.7em; transform: scale(0.85); transform-origin: top center;"}

```{r value-analysis}
#| echo: false

# Calculate value score
value_data <- data %>%
  filter(source_city == "Delhi", destination_city == "Mumbai", class == "Economy") %>%
  mutate(
    price_norm = (price - min(price)) / (max(price) - min(price) + 0.001),
    duration_norm = (duration - min(duration)) / (max(duration) - min(duration) + 0.001),
    stops_penalty = case_when(
      stops == "zero" ~ 0,
      stops == "one" ~ 0.3,
      TRUE ~ 0.6
    ),
    value_score = round((0.5 * price_norm + 0.3 * duration_norm + 0.2 * stops_penalty) * 100, 1)
  ) %>%
  arrange(value_score) %>%
  head(5)

# Display compact table
value_data %>%
  select(airline, flight, departure_time, stops, duration, price, value_score) %>%
  kable(
    col.names = c("Airline", "Flight", "Departure", "Stops", "Duration (h)", "Price (₹)", "Value Score"),
    caption = "Top 5 Best Value Flights: Delhi → Mumbai (Economy)"
  )
```

**Lower Value Score = Better Overall Value**

:::

---

## 💡 Key Insight: Value Score

::: {.callout-important}
## Smart Recommendation

The **Value Score** helps identify flights that are:

- ✅ Not necessarily the cheapest
- ✅ Not necessarily the fastest
- ✅ But the **best overall value**

**Example:** A flight costing 8% more but 25% faster has a better value score!
:::

---

# 🎯 Summary {background-color="#1E3A5F"}

---

## 📊 Dashboards Implemented

| # | Dashboard | Problem Solved |
|---|-----------|----------------|
| 1 | **Price Trends** | Best time to book |
| 2 | **Airline Comparison** | Compare airlines by value |
| 3 | **Stops Analysis** | Direct vs connecting trade-off |
| 4 | **Time Optimization** | Best departure time |
| 5 | **Route Analysis** | City-to-city optimization |
| 6 | **Value Score** | Smart flight recommendations |

---

## 🚀 Business Impact

::: {.incremental}

- 💰 **Save 15-30%** by booking at optimal times
- ⏱️ **Save 6+ hours** with informed stop decisions
- ✈️ **Better airline choices** with price/duration comparison
- ⭐ **Smarter decisions** with Value Score recommendations

:::

---

## 🙏 Thank You!

:::: {.columns}

::: {.column width="50%"}
### Technologies Used

- **R Shiny** - Interactive Dashboard
- **Plotly** - Interactive Charts
- **Tidyverse** - Data Analysis
- **Quarto** - Presentation

:::

::: {.column width="50%"}
### Next Steps

- Add real-time data integration
- Implement price prediction
- Mobile app development
- User preference learning

:::

::::

<br>

::: {.centered}
**🛫 Smart Flight Booking Dashboard**

*Making flight booking decisions data-driven*
:::
